### 2024-05-16 thesis/input_files/data_exploratory.tex
May 16th 2024

This section covers the datasets used in our experiments, in addition, attributes and features will be thoroughly described.
Next, we will explain the extract transform load pipeline (ETL) that we set up in order to ease the workflow of our comparison studies.
Finally, in order to to better understand patters within the data, we carry out an exploratory data analysis.

\section{GEFCom2014}
The first class of datasets we identified were the ones from the GEFCom 2014 competition. The data is freely hosted on Dr. Hong blog \cite{hong2016probabilistic}.
The main reason is that, these dataset are considered an a excellent test case for comparing predictive models between the EF community. Additionally, the scores of the competing models are freely available, this enables us to carry out a clean and transparent comparative study.
The GEFCom 2014 consisted of four tracks: price, load, wind power and solar power forecasting. 
The GEFCom2014 folder contains a zip file for each of the four tracks.
This thesis work is focused on providing forecasts for the load and price quantities.
\\
\subsection{GEFCom2014 Price track}
In the price forecasting track, the goal was to predict the electricity price for the next 24 hours of a single zone. 
The data can be found in the GEFCom2014-P\_V2 zip file alongside a set of instructions and the benchmark forecasts.
This data consisted of time series for the locational marginal price, for the zonal load and for the system load. The data covers the time interval ranging from the 1st of January 2011 to the 15th of June 2013, see figure \ref{fig:price_track_fig1}. 
Additionally, as the competition went on, the real observed data of the previous tasks were made available.
The fifteen target days ordered by task number are: 16/06/2013, 17/06/2013, 24/06/2013, 04/07/2013, 09/07/2013, 13/07/2013, 16/07/2013, 18/07/2013, 19/07/2013, 20/07/2013, 24/07/2013, 25/07/2013, 07/12/2013, 08/12/2013. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/price_track_fig1.png}
    \caption{Price track}
    \label{fig:price_track_fig1}
\end{figure}
\subsubsection{EDA}
In exploring this data we started off by plotting the time series of the zonal prices alongside with the total load and zonal load, see figure \ref{fig:price_track_fig1}.
After that, we plotted the scatter plot for inspecting the correlation between the predictors and the dependent variable, figure \ref{fig:gefcom_zonal_price_vs_zonal_load} and figure \ref{fig:gefcom_zonal_price_vs_total_load}. From these visualisation, we can see their goodness for predicting zonal price since there exists a clear correlation between the candidate predictors and the target variable.

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_zonal_price_vs_zonal_load.png}
    \caption{Zonal price versus zonal load}
    \label{fig:gefcom_zonal_price_vs_zonal_load}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_zonal_price_vs_total_load.png}
    \caption{Zonal price versus total load}
    \label{fig:gefcom_zonal_price_vs_total_load}
\end{figure}

@@

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_load_historical.png}
    \caption{Historical load}
    \label{fig:gefcom_load_historical}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_w_avg_historical.png}
    \caption{Historical weather}
    \label{fig:gefcom_w_avg_historical}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_load_vs_w_avg.png}
    \caption{Load versus weather}
    \label{fig:gefcom_load_vs_w_avg}
\end{figure}

\subsection{GEFCom ETL}
The provided data is raw, hence the need for a standardized extract transform and load pipeline. 
Firstly, we drop all Nans and zeros. Notice, the provided data for the challenge has a timestamp column not coherent to any time format. Thus, we first pass the data through a method extracting time information: hour, day, month and year. This info is then combined into a datetime object, we selected the ISO format.
Next, since the data does not provide the location for the target utility, the best we can do is selecting the average of weather temperatures as predictor variables. This independent variable is named w\_avg.
The same thing applies in the cleaning of the test data with the addition of an if clause for handling the different naming convention and formatting used in the last track of the challenge. The python etl pipeline has been then automated to loop every task folder through bash scripting.

@@

In the second experiment we compared various kernels in the context of load foreacsting.
For this study we considered the data of \href{https://zenodo.org/records/7907883}{SECURES-Met}. This data consists of various metereological attributes relevant for modelling electricity demand. The dataset is made up of historical data until end of 2020, while from 2021 onwards the data consists of forecasts made by a metereological model developed by the European Centre for Medium-Range Weather Forecasts (ECMWF). Notice, the forecasted data is splitted in two files according to the scenarios RCP4.5 and RCP8.5. In our experiment we considered the former. 
In this setting the independent variables available are 
direct irradiation, global radiation, power from reservoir plants, power from run of the river plants (ROR), weather temperature and wind potential.

\subsubsection{ETL}
From energy charts we downloaded the historical load along with the weather temperatures and wind speeds forecasts. Notice, the data request to the environmental variables at hourly frequency for a whole year is broken. Therefore, we downloaded each month separately and then transformed into a single csv for the whole year.

@@

\subsubsection{EDA}
Following are reported the EDA plots alongside with insights that can be extracted from them. We choose to focus the following analysis on the Swiss data only given that the plots for the German data are pretty similar.
To get started we visualized the historical load data in figure \ref{fig:CH_historical_load_2021}. As expected, electricity load exhibits a seasonal patter with highs in the winter and lows during the summer season.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_historical_load_2021.png}
    \caption{Switzerland historical load}
    \label{fig:CH_historical_load_2021}
\end{figure}
Next we built scatter plots of the independent continous variables against the load as means of visualizing correlation between them.
Figure \ref{fig:CH_load_vs_temperature_2021} plots load against temperatures while figure \ref{fig:CH_load_vs_wind_speed_2021} depicts load against wind speed. From these plots we can clearly see there exist a relationship that ties wind speeds and temperatures to load.

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_load_vs_temperature_2021.png}
    \caption{Load vs temperature Switzerland 2021}
    \label{fig:CH_load_vs_temperature_2021}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_load_vs_wind_speed_2021.png}
    \caption{Load vs wind speed Switzerland 2021}
    \label{fig:CH_load_vs_wind_speed_2021}
\end{figure}

@@

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_day_of_week_boxplot_2021.png}
    \caption{Load versus day of week boxplot}
    \label{fig:CH_day_of_week_boxplot_2021}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_is_holiday_boxplot_2021.png}
    \caption{Load versus day type boxplot}
    \label{fig:CH_is_holiday_boxplot_2021}
\end{figure}

>>>>

This section covers the datasets used in our experiments, in addition, attributes and features will be thoroughly described.
Next, we will explain the extract transform load pipeline (ETL) that we set up in order to ease the workflow of our studies.
Finally, in order to to better understand patters within the data, we carry out an exploratory data analysis.

\section{GEFCom2014}
The first class of datasets we identified were the ones from the GEFCom 2014 competition. The data is freely hosted on Dr. Hong blog \cite{hong2016probabilistic}.
The main reason is that, these dataset are considered an a excellent test case for comparing predictive models between the EF community. Additionally, the scores of the competing models are freely available, this enables us to carry out a clean and transparent comparative study.
The GEFCom 2014 consisted of four tracks: price, load, wind power and solar power forecasting. 
The GEFCom2014 folder contains a zip file for each of the four tracks.
This thesis work is focused on providing forecasts for the load and price quantities.
\\
\subsection{GEFCom2014 Price track}
In the price forecasting track, the goal was to predict the electricity price for the next 24 hours of a single zone. 
The data can be found in the GEFCom2014-P\_V2 zip file alongside a set of instructions and the benchmark forecasts.
This data consisted of time series for the locational marginal price, for the zonal load and for the system load. The data covers the time interval ranging from the 1st of January 2011 to the 15th of June 2013, see figure \ref{fig:price_track_fig1}. 
Additionally, as the competition went on, the real observed data of the previous tasks were made available.
The fifteen target days ordered by task number are: 16/06/2013, 17/06/2013, 24/06/2013, 04/07/2013, 09/07/2013, 13/07/2013, 16/07/2013, 18/07/2013, 19/07/2013, 20/07/2013, 24/07/2013, 25/07/2013, 07/12/2013, 08/12/2013. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/price_track_fig1.png}
    \caption{Price track GEFCom2014}
    \label{fig:price_track_fig1}
\end{figure}
\subsubsection{EDA}
In exploring this data we started off by plotting the time series of the zonal prices alongside with the total load and zonal load, see figure \ref{fig:price_track_fig1}.
After that, we plotted the scatter plot for inspecting the correlation between the predictors and the dependent variable, figure \ref{fig:gefcom_zonal_price_vs_zonal_load} and figure \ref{fig:gefcom_zonal_price_vs_total_load}. From these visualisation, we can see their goodness for predicting zonal price since there exists a clear correlation between the candidate predictors and the target variable.

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_zonal_price_vs_zonal_load.png}
    \caption{Zonal price versus zonal load GEFCom2014}
    \label{fig:gefcom_zonal_price_vs_zonal_load}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_zonal_price_vs_total_load.png}
    \caption{Zonal price versus total load GEFCom2014}
    \label{fig:gefcom_zonal_price_vs_total_load}
\end{figure}

@@

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_load_historical.png}
    \caption{Historical load GEFCom2014}
    \label{fig:gefcom_load_historical}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_w_avg_historical.png}
    \caption{Historical weather GEFCom2014}
    \label{fig:gefcom_w_avg_historical}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/gefcom_load_vs_w_avg.png}
    \caption{Load versus weather GEFCom2014}
    \label{fig:gefcom_load_vs_w_avg}
\end{figure}

% \subsection{GEFCom ETL}
\subsubsection{GEFCom ETL}
The provided data is raw, hence the need for a standardized extract transform and load pipeline. 
Firstly, we drop all Nans and zeros. Notice, the provided data for the challenge has a timestamp column not coherent to any time format. Thus, we first pass the data through a method extracting time information: hour, day, month and year. This info is then combined into a datetime object, we selected the ISO format.
Next, since the data does not provide the location for the target utility, the best we can do is selecting the average of weather temperatures as predictor variables. This independent variable is named w\_avg.
The same thing applies in the cleaning of the test data with the addition of an if clause for handling the different naming convention and formatting used in the last track of the challenge. The python etl pipeline has been then automated to loop every task folder through bash scripting.

@@

In the second experiment we compared various kernels in the context of load foreacsting.
For this study we considered the data of \href{https://zenodo.org/records/7907883}{SECURES-Met}. This data consists of various metereological attributes relevant for modelling electricity demand \cite{Formayer2023}. The dataset is made up of historical data until end of 2020, while from 2021 onwards the data consists of forecasts made by a metereological model developed by the European Centre for Medium-Range Weather Forecasts (ECMWF). Notice, the forecasted data is splitted in two files according to the scenarios RCP4.5 and RCP8.5. In our experiment we considered the former. 
In this setting the independent variables available are 
direct irradiation, global radiation, power from reservoir plants, power from run of the river plants (ROR), weather temperature and wind potential.

\subsubsection{ETL}
From energy charts we downloaded the historical load along with the weather temperatures and wind speeds forecasts. Notice, the data request to the environmental variables at hourly frequency for a whole year is broken. Therefore, we downloaded each month separately and then transformed them into a single csv for the whole year.

@@

\subsubsection{EDA}
Following are reported the EDA plots alongside with insights that can be extracted from them. We choose to focus the following analysis on the Swiss data only, given that the plots for the German data are pretty similar.
To get started we visualized the historical load data in figure \ref{fig:CH_historical_load_2021}. As expected, electricity load exhibits a seasonal patter with highs in the winter and lows during the summer season.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_historical_load_2021.png}
    \caption{Switzerland historical load (2021)}
    \label{fig:CH_historical_load_2021}
\end{figure}
Next we built scatter plots of the independent continous variables against the load as means of visualizing correlation between them.
Figure \ref{fig:CH_load_vs_temperature_2021} plots load against temperatures while figure \ref{fig:CH_load_vs_wind_speed_2021} depicts load against wind speed. From these plots we can clearly see there exist a relationship that ties wind speeds and temperatures to load.

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_load_vs_temperature_2021.png}
    \caption{Load vs temperature Switzerland (2021)}
    \label{fig:CH_load_vs_temperature_2021}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_load_vs_wind_speed_2021.png}
    \caption{Load vs wind speed Switzerland (2021)}
    \label{fig:CH_load_vs_wind_speed_2021}
\end{figure}

@@

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_day_of_week_boxplot_2021.png}
    \caption{Load versus day of week boxplot Switzerland (2021)}
    \label{fig:CH_day_of_week_boxplot_2021}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/CH_is_holiday_boxplot_2021.png}
    \caption{Load versus day type boxplot Switzerland (2021)}
    \label{fig:CH_is_holiday_boxplot_2021}
\end{figure}

### 2024-05-16 thesis/input_files/implementation.tex
May 16th 2024

This section is inteneded to explain and aid for reproducibilty studies. Hereafter the specific libraries used and the custom implementatios are thoroughly documented.
\\
For the list of python packages needed see the requirement.txt file on \url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.

Section documenting code
- indicate computer specifics

@@

\section{Probabilistic forecasting}
\subsection{Linear quantile regression}
The implementation of \href{https://www.statsmodels.org/dev/generated/statsmodels.regression.quantile_regression.QuantReg.html}{quantile\_regression.QuantReg} from the regression module of statsmodels has been used.
The model is fitted through iterative reweighted least squares.
\subsection{Quantile gradient boosting machine}
The implementation of \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.GradientBoostingRegressor.html}{GradientBoostingRegressor} from the sklearn.ensemble submodule has been used.
\subsection{Quantile forest}
The implementation of \href{https://pypi.org/project/quantile-forest/}{quantile\_forest.RandomForestQuantileRegressor} has been used. This estimator is compatible with scikit-learn API \cite{Johnson2024}.
\subsection{Kernel quantile regression}
Kernel quantile regression had no previously implemented open source library, thus the need of implementing our own version.
\\
The scikit-learn team provides a project template for the creation of estimators compatible with scikit-learn functionalities. Therefore, the KQR class is derived from the scikit-learn BaseEstimator and the mixin class RegressorMixin.
Our KQR class is initialized by providing a quantile, the regularization term C and the rbf kernel bandwidth gamma; the latter are the two hyperparameters of our custom estimator.
\\
In the fit method, we set up and solve the convex optimisation problem by using the cvxopt library. 
When using this library, it is important to keep two things in mind. First this library assumes the quadratic term of the optimisation problem to be multiplied by the 0.5 factor, thus we just have to provide the Q matrix with no 0.5 in front.
Secondly, in order to specify multiple inequalities we have to stack them and provide them as a unique matrix.
\\
Once a solution to the convex problem has been found, we create a mask for the support vectors of the estimator in order to estimate the constant term of our kernel quantile regressor.
\\
In the predict method, we pass a matrix $X\_eval$ of independent variables, next we compute the kernel matrix between $X\_train$ and $X\_eval$ and obtain $y\_eval$ with the formula $y=\alpha^\intercal K+b$.
\\
This estimator is compatible with useful scikit-learn in built methods like gridsearch, crossvalidation and scoring rules. Moreover, this code is compatible with sklearn in built kernel functions \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.metrics.pairwise}{sklearn.metrics.pairwise} and \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.gaussian_process.kernels}{sklearn.gaussian\_process.kernels}, functions supported by our kernel quantile regression include: gaussian rbf, absolute laplacian, mater 1/2, matern 3/2, matern 5/2, linear, cosine, sigmoid, periodic, polynomial and custom composition of kernels.
% NOTICE, these kernel implementations differ in the form of passed parameters in the former we pass in the numerator gamma while in the latter we pass l, which stands for length scale in the denominator. To cross validate overt the same range of parameter across the different kernels we made sura that gamma=1/l.
\\
Up to now, there are only two open source implementation of the quantile kernel regression. Nevertheless they are both in R, that is there exists no python, matlab or julia open source implementation.

>>>>

This section is inteneded to explain and aid for reproducibilty studies. Hereafter the specific libraries used and the custom implementatios are thoroughly documented.
\\
For the list of python packages needed see the requirement.txt file on \url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.
% - indicate computer specifics
All experiments have been carried out on a 3.2 GHz 16GB Apple M1 Pro.

Section documenting code

@@

\section{Probabilistic forecasting}
\subsection{Linear quantile regression}
The implementation of \href{https://www.statsmodels.org/dev/generated/statsmodels.regression.quantile_regression.QuantReg.html}{quantile\_regression.QuantReg} from the regression module of statsmodels has been used.
The model is fitted through iterative reweighted least squares.
\subsection{Quantile gradient boosting machine}
The implementation of \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.GradientBoostingRegressor.html}{GradientBoostingRegressor} from the sklearn.ensemble submodule has been used.
\subsection{Quantile forest}
The implementation of \href{https://pypi.org/project/quantile-forest/}{quantile\_forest.RandomForestQuantileRegressor} has been used. This estimator is compatible with scikit-learn API \cite{Johnson2024}.
\subsection{Kernel quantile regression}
Kernel quantile regression had no previously implemented python open source library, thus the need of implementing our own version.
\\
The scikit-learn team provides a project template for the creation of estimators compatible with scikit-learn functionalities. Therefore, the KQR class is derived from the scikit-learn BaseEstimator and the mixin class RegressorMixin.
Our KQR class is initialized by providing a quantile, the regularization term C, the kernel family and its corresponding hyperparameters.
\\
In the fit method, we set up and solve the convex optimisation problem by using the cvxopt library, see offical manual \cite{vandenberghe2010cvxopt} for a reference.
When using this library, it is important to keep two things in mind. First this library assumes the quadratic term of the optimisation problem to be multiplied by the 0.5 factor, thus we just have to provide the Q matrix with no 0.5 in front.
Secondly, in order to specify multiple inequalities we have to stack them and provide them as a unique matrix.
\\
Once a solution to the convex problem has been found, we create a mask for the support vectors of the estimator in order to estimate the constant term of our kernel quantile regressor.
\\
In the predict method, we pass a matrix $X\_eval$ of independent variables, next we compute the kernel matrix between $X\_train$ and $X\_eval$ and obtain $y\_eval$ with the formula $y=\alpha^\intercal K+b$.
\\
This estimator is compatible with useful scikit-learn in built methods like gridsearch, crossvalidation and scoring rules. Moreover, this code is compatible with sklearn in built kernel functions \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.metrics.pairwise}{sklearn.metrics.pairwise} and \href{https://scikit-learn.org/stable/modules/classes.html#module-sklearn.gaussian_process.kernels}{sklearn.gaussian\_process.kernels}, functions supported by our kernel quantile regression include: gaussian rbf, absolute laplacian, mater 1/2, matern 3/2, matern 5/2, linear, cosine, sigmoid, periodic, polynomial and custom composition of kernels.
% NOTICE, these kernel implementations differ in the form of passed parameters in the former we pass in the numerator gamma while in the latter we pass l, which stands for length scale in the denominator. To cross validate overt the same range of parameter across the different kernels we made sura that gamma=1/l.

% \\
% Up to now, there are only two open source implementation of the quantile kernel regression. Nevertheless they are both in R, that is there exists no python, matlab or julia open source implementation.
